Windowed residual-minimization model reduction for dynamical systems

### 2019-10-03 main.tex
into 2.2.2

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG ROMs; Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
formulation; Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods; Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs;
Section~\ref{sec:numerical_experiments} presents numerical experiments;
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with stylized bold text (i.e., $\boldsymbol
x$), vectors with standard bold text (i.e., $\mathbf{x}$), 
matrices with capital bold letters (i.e., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic fonts (i.e., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical setting}\label{sec:math}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \frac{d\stateFOM}{dt} = \velocity(\stateFOM,t), \qquad \stateFOM(0) = \stateFOMIC,
\end{equation}
where $t \in [0,T]$ denotes time with 
$T \in \mathbb{R}^+$ denoting the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial conditions,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Projection-based reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{\spatialAcronym\ trial subspaces}}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection)
	\item \textit{\spaceTimeAcronym\ trial subspaces}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection. 
	\KTC{HERE}

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution, $\stateFOM$, with an approximate solution, $\approxstate$, that lives in a $K$-dimensional affine trial subspace, 
\begin{align}\label{eq:spatial_subspace}
& \stspaceS \defeq \trialspace \otimes \timeSpace,\\ 
& \trialspace \defeq \text{Range}(\basismat) +  \stateIntercept,  \label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, \label{eq:spatial_subspace3} 
\end{align}
where we refer to $\basismat \in \RRStar{\romdim}{\fomdim}$, with $\RRStar{\romdim}{\fomdim}$ denoting the set  $\{ \mathbf{X} \in \RR{\fomdim \times \romdim} \big| \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$, as the spatial trial basis matrix, $\trialspace$ as the spatial trial subspace, and $\stateIntercept \in \mathbb{R}^{\fomdim}$ as the reference state. 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$ are typically constructed through, e.g., POD~\cite{berkooz_turbulence_pod}, the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}, etc. 
ROMs leveraging \spatialAcronym\ trial subspaces approximate the state as,
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) \defeq \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$, with $\genstate : [0,T] \rightarrow \mathbb{R}^{\romdim}$, are the generalized coordinates. Substituting the approximation given in~\eqref{eq:affine_trialspace} into~\eqref{eq:FOM} and performing orthogonal $\elltwo$ projection of the initial conditions onto the trial subspace, one obtains,
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d }{dt}\genstateArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Equation~\eqref{eq:g_truncation} provides an over-determined system of $N$ equations for the $K$ generalized coordinates. Dimension reduction is predominantly performed with either the \textit{Galerkin method} or the \textit{Least-Squares Petrov--Galerkin method} (LSPG). We now outline these two methods.

@@

The Galerkin method is (traditionally) obtained by the following steps: (1.) The approximation~\eqref{eq:affine_trialspace} is substituted into~\eqref{eq:FOM}, and (2.) the resulting residual is enforced to be $\stweightingMatArg{}$-orthogonal to the trial subspace, with $\stweightingMatArg{ } \equiv \stweightingMatOneArg{ }^T \stweightingMatOneArg{ }$ being a positive (semi-)definite weighting matrix (commonly taken to be $\mathbf{I}$). This projection process results in what is known as the Galerkin method,

@@

\frac{d}{dt}\genstateGalerkinArg{}{t} = \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace
\genstateGalerkinArg{}{t} + \stateIntercept,t), \qquad \genstateGalerkin(0) = \genstateIC, \qquad t \in [0,T],

@@

where $\massArgnt{ } \equiv \basisspace^T \stweightingMatArg{ } \basisspace$ is a mass matrix. 
Model reduction via the Galerkin method is seen to result in a $K$-dimensional dynamical system for the 
generalized coordinates. 
The Galerkin method can alternatively be viewed as a residual minimization method~\cite{carlberg_lspg_v_galerkin}. In this setting, the Galerkin ROM can be 
written as,
\frac{d}{dt}\genstateGalerkinArg{}{t} = \underset{ \genstateyDiscrete \in \RR{\romdim}  }{\text{arg\,min}} || \basisspace \hat{\stateyDiscrete} -  \velocity(\basisspace \genstateGalerkinArg{}{t} + \stateIntercept,t) ||_{\stweightingMatArg{n}}^2. 

@@

@@

The Galerkin ROM corresponds to \textit{time-instantaneous} minimization of the FOM ODE residual.

@@

The Galerkin method can be unstable and inaccurate for non-self-adjoint and nonlinear problems. 
An alternative model-reduction approach that enjoys several advantageous over the Galerkin method is the Least-Squares Petrov--Galerkin (LSPG) method~\cite{carlberg_lspg_v_galerkin,carlberg_thesis,carlberg_gnat,bui_unsteady,bui_thesis}. LSPG is defined at the fully discrete level and minimizes the fully discrete FOM O$\Delta$E residual (i.e., the FOM ODE residual after temporal discretization) at each time-step instance.  We outline the LSPG method for linear multistep methods as a demonstrative example. Defining a discretization of the continuous time domain $[0,T]$ into $N_t$ time instances $t^n = n\Delta t, n=1,2,\ldots,N_t$ with a constant time step $\Delta t = T/N_t$, temporal discretization via a linear multistep method yields the FOM O$\Delta$E,

>>>>

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG ROMs; Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
formulation; Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods; Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs;
Section~\ref{sec:numerical_experiments} presents numerical experiments;
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with stylized bold text (i.e., $\boldsymbol
x$), vectors with standard bold text (i.e., $\mathbf{x}$), 
matrices with capital bold letters (i.e., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic fonts (i.e., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical setting}\label{sec:math}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \frac{d\stateFOM}{dt} = \velocity(\stateFOM,t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{Projection-based reduced-order models}
Projection-based ROMs generate approximate solutions to the FOM
	ODE~\eqref{eq:FOM} by approximating the state in a low-dimensional trial
	subspace. For dynamical 
	systems, two distinct types of space--time trial subspaces are commonly used: 
\begin{enumerate} 
	\item \textit{\spatialAcronym\ trial subspaces}. These 
	trial subspaces reduce only the spatial dimension of the FOM, and are commonly used
		in model-reduction methods based on spatial projection followed by time
		integration (e.g., Galerkin projection, LSPG projection)
	\item \textit{\spaceTimeAcronym\ trial subspaces}. These 
	trial subspaces reduce both the spatial and temporal dimensions of the FOM,
		and are commonly employed 
		employed in model-reduction methods based on space--time projection (e.g.,
		space--time LSPG projection).
\end{enumerate}
 We now describe these two types of trial subspaces and their application to 
	Galerkin, LSPG, and space--time LSPG projection. 
	\KTC{HERE}

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{align}\label{eq:spatial_subspace}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, \label{eq:spatial_subspace3} 
\end{align}
where 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain
	$[0,T]$, $\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} in
Eq.~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d\genstate }{dt} = \velocity(\basisspace
\genstate + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

Galerkin projection reduces the number of equations in
Eq.~\eqref{eq:g_truncation} by enforcing orthogonality of the associated
residual to the trial subspace in the (semi-)inner product induced by the 
positive (semi-)definite matrix $\stweightingMatArg{}\equiv \stweightingMatOneArg{ }^T
\stweightingMatOneArg{ }$ (commonly set to
$\stweightingMatArg{}=\mathbf{I}$), i.e.,

@@

\frac{d\genstateGalerkin}{dt} = \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace
\genstateGalerkin + \stateIntercept,t), \qquad \genstateGalerkin(0) = \genstateIC, \qquad t \in [0,T],

@@

where $\massArgnt{ } \equiv \basisspace^T \stweightingMatArg{ } \basisspace$
denotes a mass matrix.  Note that the Galerkin ODE~\eqref{eq:g_truncation2} is
a system of $K$ equations in $K$ unknowns.
As demonstrated in Ref.~\cite{carlberg_lspg_v_galerkin}, Galerkin projection
can be viewed alternatively as a residual-minimization
method. In particular, the Galerkin ODE~\eqref{eq:g_truncation2} is equivalent
to
\begin{equation}\label{eq:GalOptimal}
\frac{d\genstateGalerkin}{dt} = \underset{ \genstateyDiscrete \in \RR{\romdim}
	}{\text{arg\,min}} \| \basisspace \hat{\stateyDiscrete} -
	\velocity(\basisspace \genstateGalerkin + \stateIntercept,t)
	\|_{\stweightingMatArg{n}}^2,

@@

where $\|\mathbf{x}\|_{\stweightingMatArg{n}}\equiv
\sqrt{\mathbf{x}^T\stweightingMatArg{n}\mathbf{x}}$.

@@

Thus, velocity $\frac{d\genstateGalerkin}{dt}(t)$ computed by Galerkin projection minimizes the
time-instantaneous FOM ODE residual evaluated at the state $\basisspace
\genstateGalerkin(t) + \stateIntercept$ and time $t$.

@@

Despite its time-instantaneous optimality property expressed in
Eq.~\eqref{eq:GalOptimal}, Galerkin projection can yield unstable and
inaccurate trajectories, particularly if the velocity is not self-adjoint or
is nonlinear. 
Least-squares Petrov--Galerkin (LSPG)
projection~\cite{carlberg_lspg_v_galerkin,carlberg_thesis,carlberg_gnat,bui_unsteady,bui_thesis}
is an alternative approach that exhibits several advantages over Galerkin
projection.
Rather than minimize the (time-continuous) FOM ODE residual as in Galerkin
projection, LSPG minimizes
the residual arising at a given time step after applying time discretization to the FOM ODE; we
refer to the residual arising at the fully discrete level as the FOM O$\Delta$E residual.  
We now describe LSPG projection when a linear multistep method is employed for
time discretization; Ref.~\cite{carlberg_lspg_v_galerkin} derives LSPG projection for Runge--Kutta
schemes as well. 
Defining a discretization
of the continuous time domain $[0,T]$ into $N_t$ time instances $t^n = n\Delta
t, n=1,2,\ldots,N_t$ with a constant time step $\Delta t = T/N_t$, temporal
discretization via a linear multistep method yields the FOM O$\Delta$E,
